---
title: 'Practical Machine Learning : Prediction of excercise manner'
output: html_document
---

Background
==========
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset). 


Data 
====
The training data for this project are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment. 

Goal
====
The goal of this project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. Following report describes how model is built, how cross validation is used, what is the expected out of sample error is, and why the choices made. The prediction model is used to predict 20 different test cases

Load the required paackages
===========================
```{r LoadPackages}
library(caret)
library(randomForest)
```

Getting and Cleaning Data
=========================
1. Download data sets using provided URLs if not exist.
```{r DownloadData}
# Download training data set if not exists
trainDS <- "pml-training.csv"
if (!file.exists(trainDS)) {
  download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", 
                destfile=trainDS, method="curl")
}

# Download test data set if not exists
testDS  <- "pml-testing.csv"
if (!file.exists(testDS)) {
  download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv", 
                destfile=testDS, method="curl")
}
```

2. Read data sets into data frames.
```{r ReadData}
# Read training data set into a data frame
trainDF <- read.csv(trainDS)

# Display the dimensions of training data frame
dim(trainDF)

# Read test data set into a data frame
testDF <- read.csv(testDS)

# Display the dimensions of test data frame
dim(testDF)
```

3. Clean data frames to remove the columns containing missing values and columns that cannot be used for measurements.
```{r CleanData}
# Remove the columns that contain missing values
trainDF <- trainDF[, colSums(is.na(trainDF)) == 0] 
testDF <- testDF[, colSums(is.na(testDF)) == 0] 

# Remove non numeric columns which cannot be used for measurements except classe
trainDFclasse <- trainDF$classe
trainDF <- trainDF[, sapply(trainDF, is.numeric)]
trainDF$classe <- trainDFclasse
testDF <- testDF[, sapply(testDF, is.numeric)]

# Remove first 4 columns which are not used for measurements
trainDF <- trainDF[, 5:length(names(trainDF))]
testing <- testDF[, 5:length(names(testDF))]

# Display the dimensions of training data frame
dim(trainDF)

# Display the dimensions of testing data frame
dim(testing)
```

4. Partition the train data frame into a pure train data frame(70%) and a validation data frame(30%) used for cross validation.
```{r PartitionData}
# Set seed for reproducibility
set.seed(13234)

# Partition train data frame into 70 : 30 data frames
inTrain = createDataPartition(trainDF$classe, p = 0.7, list=FALSE)
training = trainDF[inTrain,]
validation = trainDF[-inTrain,]

# Display the dimensions of training data frame 
dim(training)

# Display the dimensions of validation data frame
dim(validation)

# Display the dimensions of testing data frame
dim(testing)
```

Building Model
==============
We will build model for training data using method random forest and using train control method 5-fold cross validation. We have used method random forest as it results into high accuracy rate. 
```{r BuildModel}
# Train the model
modFit <- train(classe ~., 
                method="rf", 
                data=training, 
                trControl= trainControl(method='cv', 5),
                ntree=250)

# Display the model
modFit

## Display the variable importance
varImp(modFit)

# Display the final model
modFit$finalModel
```

Evaluating Model for validation data set
========================================
We will evaluate the model on validation data set.
```{r EvaludateModelValidation}
# Evaluate model on validation data set
predictValidation <- predict(modFit, validation)

# Display confusionMatrix
cm <- confusionMatrix(validation$classe, predictValidation)
cm

# Estimate the accuracy percentage
accuracy <- as.numeric(cm$overall[1]) * 100
accuracy

# Estimate out of sample error percentage
out_of_sample_error <- (1 - as.numeric(cm$overall[1]))* 100
out_of_sample_error
```

So, the estimated accuracy of the model is `r accuracy`% and the estimated out of sample error is `r out_of_sample_error`%.

Evaluating Model for test data set
==================================
We will evaluate the model on test data set.
```{r EvaludateModelTest}
## Evaluate model on testing data set
predict(modFit, testing)
```

Figures
=======
We will plot various models.
```{r PlotModel}
# plot model 
plot(modFit)

# plot the final model
plot(modFit$finalModel)

# plot the variable importance 
varImpPlot(modFit$finalModel)
```